Concurrent Programming with CXL

### 2020-05-19 book/paper.tex
Example of using methods and pointers
Latest version

\documentclass{report}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}

@@

CXL methods look and feel much the same as Python methods, but there
are some differences you should be aware of.
First, as already noted, there is no \texttt{return} statement.
A CXL method returns when the last statement in the method has
completed.

@@

A method \texttt{m} with argument \texttt{a} is invoked in its
most basic form as follows (assigning the result to \texttt{x}).
\begin{code}
\begin{verbatim}
var x = m a;
\end{verbatim}
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
as \texttt{a}.
If you invoke \texttt{m()}, the argument is \texttt{()},
which is the empty tuple.
If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
the tuple consisting of values \texttt{a} and \texttt{b}.
You have already seen examples of this: the expression
\texttt{atLabel.cs} invokes method \texttt{atLabel} with
argument \texttt{.cs}.

You may note that all this looks familiar.  Indeed, the syntax
is the same as that for dictionaries (see Chapter~\ref{ch:cxlvalues}).
Both dictionaries and methods map CXL values to CXL values,
and their syntax is indistinguishable.
If \texttt{f} is either a method or a
dictionary, and \texttt{x} is an arbitrary CXL value, then
then \texttt{f x}, \texttt{f(x)}, and \texttt{f[x]} are all
the same expression in CXL.

If you want to invoke a method but are not interested in its result,
you have to use a \texttt{call} statement like so:
\begin{code}
\begin{verbatim}
call m a;
\end{verbatim}
\end{code}

\chapter{Spinlock}
Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual exclusion.

>>>>

\documentclass{report}
\usepackage{fullpage}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{tcolorbox}
\usepackage{verbatim}
\usepackage{comment}

@@

A method \texttt{m} with argument \texttt{a} is invoked in its
most basic form as follows (assigning the result to \texttt{r}).
\begin{code}
\begin{verbatim}
r = m a;
\end{verbatim}
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
as \texttt{a}.
If you invoke \texttt{m()}, the argument is \texttt{()},
which is the empty tuple.
If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
the tuple consisting of values \texttt{a} and \texttt{b}.
You have already seen examples of this: the expression
\texttt{atLabel.cs} invokes method \texttt{atLabel} with
argument \texttt{.cs}.

You may note that all this looks familiar.  Indeed, the syntax
is the same as that for dictionaries (see Chapter~\ref{ch:cxlvalues}).
Both dictionaries and methods map CXL values to CXL values,
and their syntax is indistinguishable.
If \texttt{f} is either a method or a
dictionary, and \texttt{x} is an arbitrary CXL value, then
\texttt{f x}, \texttt{f(x)}, and \texttt{f[x]} are all
the same expression in CXL.

\begin{figure}
\begin{code}
\verbatiminput{../PetersonMethod.cxl}
\end{code}
\caption{Peterson's Algorithm accessed through methods.}
\label{fig:petersonmethods}
\end{figure}

It is often the case that methods that update shared variables through
pointers do not actually have to return a result to be useful.
In CXL, if you want to invoke a method that does not return a result
or if you are simply not interested in it,
you have to use a \texttt{call} statement like so:
\begin{code}
\begin{verbatim}
call m a;
\end{verbatim}
\end{code}

Figure~\ref{fig:petersonmethods} again shows Peterson's algorithm,
but this time with methods defined to enter and exit the critical
section.
You can put the first three methods in its own CXL source file
and include it using the CXL \texttt{import} statement.  This would
make the code re-usable by other applications.  For those who are
extra adventurous: you can add the \texttt{P\_enter} and
\texttt{P\_exit} methods to the \texttt{P\_mutex} dictionary
like so:
\begin{code}
\begin{verbatim}
dict{ .turn: 0, .flags: [ False, False ], .enter: P_enter, .exit: P_exit }
\end{verbatim}
\end{code}
That would allow you to simulate object methods.

\chapter{Spinlock}

\begin{figure}
\begin{code}
\verbatiminput{UpTAS.cxl}
\end{code}
\caption{Mutual Exclusion using a ``spinlock'' based on test-and-set.}
\label{fig:tas}
\end{figure}

Figure~\ref{fig:uplock} showed a faulty attempt at solving mutual
exclusion using a lock.  The problem with the implementation of the
lock is that checking the lock and setting it if it is available is
not \emph{atomic}.  Thus multiple processes contending for the lock
can all ``grab the lock'' at the same time.  While Peterson's
algorithm gets around the problem, it is not efficient, especially
if generalized to multiple processes.  Instead, processors provide
so-called \emph{interlock instructions}: special machine instructions
that can read memory and then write it in an indivisible step.

While the CXL machine does not have any built-in interlock instructions,
it does have support for executing multiple instructions atomically.
This feature is available in the CXL language through its \texttt{atomic}
statement.
You already saw an example of this in Figure~\ref{fig:petersonproof}.
We can use \texttt{atomic} blocks to implement a wide variety of
interlock operations.
For example, we could fix the program in Figure~\ref{fig:inc} by
constructing an atomic increment operation for a counter, like so:
\begin{code}
\begin{verbatim}
def atomic_inc(ptr):
    atomic:
        ^ptr = ^ptr + 1;
    ;
;
count = 0;
call atomic_inc(&(count));
\end{verbatim}
\end{code}

Many CPUs have an atomic ``test-and-set'' operation.
Method \texttt{tas} in Figure~\ref{fig:tas} shows its specification.
Here \texttt{lk} points to a shared variable holding a
boolean.  The operation assigns \texttt{True} to the
variable but returns the value the variable had \emph{before}
the assignment.
Figure~\ref{fig:tas} goes on to show how the program
in Figure~\ref{fig:uplock} can be made correct.
Using \texttt{tas} a process can wait until \texttt{lock} is
\texttt{False} and then set it to \texttt{True} before another
process is able to see its old value.

Compared to Peterson's Algorithm, this approach is general for
any number of processes.  To make sure, you may want to try it
for three or more processes.  In fact, you may want to replace
the last two lines by:
\begin{code}
\begin{verbatim}
for i in 1..5:
    spawn process(), ();
;
\end{verbatim}
\end{code}
This creates 5 processes with the same name tag.  CXL can
exploit their symmetry to do a much more efficient state
search.  Try replacing it with the following code and see
the difference:
\begin{code}
\begin{verbatim}
for i in 1..5:
    spawn process(), i;
;
\end{verbatim}
\end{code}

\chapter{Local Variables}

@@
